\textbf{ Matching the forecast horizon with the relevant ecological processes }

### 2019-05-24 doc/manuscript.tex
Finished this draft

No such consensus is available for exactly how to tackle this challenge.
Ecologists rely on many different approaches and models for making predictions, and 
frequently argue about the virtues of process-based vs. phenomenological models (refs?).
We often justify investments in basic research by claiming that improved process-level understanding will improve predictions (e.g., Adler et al. 2012),
and it is easy to find examples of how extrapolation of linear relationships leads to poor predictions under novel conditions (refs?).
But in many cases simple phenomenological models make better predictions than complex process-based models with high parameter
uncertainty (refs? \citealt{tredennick_we_2017}?).

The pragmatic solution to these debates is clear: we should forecast using the models that make the best predictions.
Identifying these models requires rigorously testing predictions against observations. However, many common ecological predictions,
such as a species' geographic range in the year 2100, are virtually impossible to test (but see paleo hindcasting refs,Williams refs? Blois?), 
leading \citep{dietze_iterative_2018} to call for greater emphasis on near-term forecasting. The assumption is that building 
our capacity to make short-term forecasts will ultimately improve our long-term forecasts.

@@

We make two forecasts based on two different phenomenological models, each representing processes operating at different time scales. 
The first phenomenological model represents what we call a ``spatial approach.'' We correlate the mean environment with the mean of an ecological state or rate 
 across many sites. This is the approach commonly used to predict population distribution or abundance as a function of climate (SDM refs) or
 primary production as a function of mean precipitation (Lauenroth refs). The approach is attractive because it addresses broad spatial scales relevant to 
 management, and because the necessary data are widely available and statistical analyses are straightforward. The approach also has well-known weaknesses.
Criticism often focuses on the assumption that current spatial patterns are in equilibrium with climate (refs?). Another issue is that these models are not dynamic; 
they provide no information about how quickly the system will move from the current state to the predicted, future state. Because the patterns in the data have developed 
over long time spans, they are unlikely to provide much information about dynamic changes on the order of years to decades. Moreover, transient dynamics could 
prevent the system from ever reaching the predicted steady state (Urban ref?).

The second model represents what we call a ``temporal approach.'' We correlate interannual variation in an ecological response with interannual variation in the environment, 
typically at just one site. This approach is often used to study population or vital rate fluctuations as a function of weather \citep{dalgleish_climate_2011}, 
or primary production as a function of precipitation \citep{lauenroth_long-term_1992}. The advantages of this approach include is focus on temporal dynamics, meaning it 
can predict rates of change, and validation of the resulting near-term predictions are straightforward. However, models built using this approach typically cover 
a limited spatial extent (but see \cite{kleinhesselink_response_2018}, others), and ignore slower processes, such as evolutionary adaptation or turnover in community composition, 
that could alter the underlying process at longer time scales.

We compare forecasts from both models to the true, simulated dynamics to show that responses to environmental change reflect both kinds of processes, but at different time scales. To quantify this pattern, we combine forecasts from the spatial and temporal approaches with a weighted average and then show how the optimal model weights change over time: we expect the temporal model to best predict short-term dynamics, and the spatial model to best predict long-term dynamics. How rapidly that transition occurs depends on the rate of  ``slow" processes such as evolution and species turnover.

@@

A common problem for conservation biologists and natural resource managers is anticipating the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just the focal species, 
skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do
altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by Alexander et al. (ref). 
The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient.
Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: 
cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by 
fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. 
Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. All computer code is available on Github: \texttt{https://github.com/pbadler/space-time-forecast})

@@

To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix \ref{models} for full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-weights-spp}) 
because the weights were determined by fitting directly to those data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from from being dominated by the interannual 
variations captured in the temporal model to being dominated by the steady-state equilibrium captured by the spatial model.
The most rapid shifts in model weights occur during the period when warm-adapted, faster growing species are increasing most rapidly
in abundance (Fig. \ref{fig:community-weights-spp}). A true, combined forecast would require a method to determine the model weights \emph{a priori}.

@@

The compositional turnover affecting our focal species also influences total biomass, linking community and ecosystem dynamics. We repeated our focal species analysis 
for total community biomass, and the results were similar: the temporal model initially made the best forecasts immediately following the onset of the temperature
increase, but as the identify and abundances of species at the study site changed, the model weights rapidly shifted to the spatial model (Figs. \ref{fig:community-models-total} - \ref{fig:community-weights-total}).

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location (refs). Like the shifts
in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales.
We demonstrate this process with a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A; the full model description is provided in Appendix \ref{eco-evo}, and all computer code is available on Github: \texttt{https://github.com/pbadler/space-time-forecast}). Genotype frequencies will shift across a gradient of mean annual temperature: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B).

>>>>

No such consensus is available for exactly how to tackle this challenge.
Ecologists rely on many different approaches and models for making predictions, and 
frequently argue about the virtues of process-based vs. phenomenological models (refs?).
We often justify investments in basic research by claiming that improved process-level understanding will improve predictions (e.g.,\citealt{Adler2012}),
and it is easy to find examples of how extrapolation of linear relationships leads to poor predictions under novel conditions \citep{Fitzpatrick,Veloz2012}.
But in many cases simple phenomenological models make better predictions than complex process-based models with high parameter
uncertainty (refs? \citealt{tredennick_we_2017}?).

The pragmatic solution to these debates is clear: we should forecast using the models that make the best predictions.
Identifying these models requires rigorously testing predictions against observations. However, many common ecological predictions,
such as a species' geographic range in the year 2100, are virtually impossible to test (but see \citealt{Maguire2015}), 
leading \citep{dietze_iterative_2018} to call for greater emphasis on near-term forecasting. The assumption is that building 
our capacity to make short-term forecasts will ultimately improve our long-term forecasts.

@@

We make two forecasts based on two different phenomenological models, each representing processes operating at different time scales. 
The first phenomenological model represents what we call a ``spatial approach.'' We correlate the mean environment with the mean of an ecological state or rate 
 across many sites. This is the approach commonly used to predict population distribution or abundance as a function of climate \citep{elith_species_2009} or
 primary production as a function of mean precipitation \citep{Sala1988}. The approach is attractive because it addresses broad spatial scales relevant to 
 management, and because the necessary data are widely available and statistical analyses are straightforward. The approach also has well-known weaknesses.
Criticism often focuses on the assumption that current spatial patterns are in equilibrium with climate (refs?). Another issue is that these models are not dynamic; 
they provide no information about how quickly the system will move from the current state to the predicted, future state. Because the patterns in the data have developed 
over long time spans, they are unlikely to provide much information about dynamic changes on the order of years to decades. Moreover, transient dynamics could 
prevent the system from ever reaching the predicted steady state \citep{Urban2012}.

The second model represents what we call a ``temporal approach.'' We correlate interannual variation in an ecological response with interannual variation in the environment, 
typically at just one site. This approach is often used to study population or vital rate fluctuations as a function of weather \citep{dalgleish_climate_2011}, 
or primary production as a function of precipitation \citep{lauenroth_long-term_1992}. The advantages of this approach include is focus on temporal dynamics, meaning it 
can predict rates of change, and validation of the resulting near-term predictions are straightforward. However, models built using this approach typically cover 
a limited spatial extent (but see \citealt{kleinhesselink_response_2018}, others), and ignore slower processes, such as evolutionary adaptation or turnover in community composition, 
that could alter the underlying process at longer time scales.

We compare forecasts from both models to the true, simulated dynamics to show that responses to environmental change reflect both kinds of processes, but at different time scales. To quantify this pattern, we combine forecasts from the spatial and temporal approaches with a weighted average and then show how the optimal model weights change over time: we expect the temporal model to best predict short-term dynamics, and the spatial model to best predict long-term dynamics. How rapidly that transition occurs depends on the rate of  ``slow" processes such as species turnover and evolution.

@@

A common problem for conservation biologists and natural resource managers is anticipating the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just the focal species, 
skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do
altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018}. 
The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient.
Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: 
cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by 
fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. 
Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. All computer code is available on Github: \texttt{https://github.com/pbadler/space-time-forecast}).

@@

To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix \ref{models} for full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}) 
because the weights were determined by fitting directly to those data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from from being dominated by the interannual 
variations captured in the temporal model to being dominated by the steady-state equilibrium captured by the spatial model.
The most rapid shifts in model weights occur during the period when warm-adapted, faster growing species are increasing most rapidly
in abundance (Fig. \ref{fig:community-weights-spp}). A true, combined forecast would require a method to determine the model weights \emph{a priori}.

@@

The compositional turnover affecting our focal species also influences total biomass, linking community and ecosystem dynamics. We repeated our focal species analysis 
for total community biomass, and the results were similar: the temporal model initially made the best forecasts immediately following the onset of the temperature
increase, but as the identify and abundances of species at the study site changed, the model weights rapidly shifted to the spatial model (Figs. \ref{fig:community-models-total} to \ref{fig:community-weights-total}).

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location (refs). Like the shifts
in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales.
We demonstrate this process with a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). The full model description is provided in Appendix \ref{eco-evo}, and all computer code is available on Github  (\texttt{https://github.com/pbadler/space-time-forecast}). Genotype frequencies will shift across a gradient of mean annual temperature: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B).
